Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2020-10-01 manuscript.tex
Updates from Overleaf

%\newcommand*{\rood}[1]{\textcolor{red}{#1}}
\newcommand*{\rood}[1]{#1}
%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
\newcommand*{\blauw}[1]{#1}
%\newcommand*{\groen}[1]{\textcolor{green}{#1}}
\newcommand*{\groen}[1]{#1}
%\newcommand*{\blauwr}[1]{\textcolor{blue}{#1}}
\newcommand*{\blauwr}[1]{#1}
\newcommand{\Expect}[1]{\mathrm{E}\left[#1\right]}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

@@

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems.pdf, Rothemund2006FoldingPatterns, Gu2010ALine}. Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales, and dynamic DNA nanotechnology has been used to perform basic calculation and to probe single molecules via temporal DNA signatures\citep{MhatreV.HoJi-AnnLee2012NIHAccess, Bui2018, Shah2019}. Both technologies rely on the hybridization reaction between complementary DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions. Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble meta-stable states facilitates the transition. Recent breakthrough studies have coupled experimental techniques with machine learning and MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}. Where these studies focus on association and dissociation kinetics alone, we broaden our analysis into higher order dynamical processes and meta-stable intermediates. The stability of certain intermediates states, such as out-of-register or shifted base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014, Xiao2019, Maciejczyk2014DNAModel}). Furthermore, frayed structures and dynamics have been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}. Sandstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability of frayed states was dictated by G:C base placement 10-mer oligonucleotide sequences \citep{Sanstead2016}. In this work, we study the same four sequences explored by Sandstead et al in an effort to uncover the sequence-dependent dynamics and their relation to metastable structures mentioned above.

Our understanding of DNA dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, or other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these present technical difficulties and are hampered by data collection rates \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices}. Förster resonance energy transfer (FRET) analysis, particularly when coupled with methods mentioned above, provides additional resolution, but it is unclear how fluorescent tags may interfere with the dynamics of short nucleotides \citep{Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several high-level computational models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}. Although these models provide experimentally verified speed-ups compared to all atom simulations, the long timescales on which DNA hybridization and dissociation events occur make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}.Other computational works use dramatically elevated temperature or denaturing solvent concentrations induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied certain aspect the overall dynamics process in one direction.

\textcolor{purple}{Could add more background on here about all-atom simulations, but these rarely explore fully hybridization/dissocition events}

In this work, we use the coarse-grain 3 Sites Per Nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.  We leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization  \citep{Jin2019, Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}. We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}. Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstates clustering can be performed. We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).

We find that GC base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior. By evaluating equilibrium trajectories we can study the relevance of meta-stable states during both the hybridization and dissociation process. Furthermore we can perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another. Because SRVs generate an optimized low dimensional basis, we show that we can access higher resolution MSMs (shorter lag time) and generate more detailed models. Additionally, we can compare slow dynamical modes and meta-stable states between sequence-specific SRV-MSMs. Within these meta-stable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes. Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant DNA dynamics \textcolor{purple}{Not sure how much to say about this here, but I think these experiments should be mentioned in the intro}. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequences dependent dynamics, meta-stable structures, and relative timescales.

@@

\subsection{\label{sec:methods}3spn2 Model}

\textcolor{purple}{Is some additional information warranted on the model itself before going into this specific setup?}

@@

We initialized four sequences previously investigated by Sandstead et al. -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands according to 3SPN.2 documentation \citep{Sanstead2016, Phys2014}. We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}. In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end-to-end extension length of a single strand and the force cutoff (using Ewald summation method is set at 20 A). This translated to a box size of 77.74 Å and an effective oligo concentration of 7 mM. We used an Ewald potential to calculate long range Coulombic interaction between DNA and ions. We used a Debye-Huckel screening potential to account for phosphate-backbone interactions and preserve the persistence length and intrinsic curvature of DNA while taking the effects of ion-DNA interactions\citep{Hinckley2015}. We ran our simulations in the NVT ensemble and fix temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}. We set the simulation temperature was set about 1 K higher than empirical sequence-specific melting temperature in order to maximize transitions between dissociated and hybridized states. We use a 15 fs time step and $3*10^{8}$ steps in each simulation, equating to 4.5$\mu$s simulation time. Base pair coordinates were saved every 30 ps, and full frames including backbones and ions were saved every 3 ns for visualization purposes. For each sequences, 100 simulations were performed in parallel, consuming about 32 serial cpu-hours of computation time for each independent simulation. An approximate Tmelt Boltzman distribution was replicated by initializing half of runs from the hybridized state and half from a random dissociated state. In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.

@@

All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}. Based on the self-complementary nature of each sequence, we averaged permutable distances (45 pairs in total) together. This permutation reduction follows a similar procedure used in TICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}. The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}. The VAMP-2 score uses the covariances of a set of inputs estimates the transfer operator of a dynamical system (equation below), providing a robust and object means to evaluate various parameters and models. In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$. The score $R_2$ can then be found for $\chi$ by applying the VAMP principle with cross-validation.

\begin{align*}
 	\mathscr{C}_{00}=&\Expect{\chi(x_t)\chi(x_t)^\intercal}_t\\
 	\mathscr{C}_{01}=&\Expect{\chi(x_t)\chi(x_{t+\tau})^\intercal}_t\\
 	\mathscr{C}_{11}=&\Expect{\chi(x_{t+\tau})\chi(x_{t+\tau})^\intercal}_{t+\tau}\\
	\\
 	R_2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
 	%R_2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
\end{align*}\label{CK1}

@@

SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The framework uses a twin-lobed artificial neural network, similar in structure to the network employed in VAMPnets \citep{Mardt2018VAMPnetsKinetics}, to learn a low dimensional representation of input features. This representation is optimized for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated. The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs. Other methods, such as time-lagged independent component analysis (TICA) and kernel TICA (kTICA), have been employed in place of SRVs but suffer limitations from the linear transformation in the case of the former and high computational cost in the case of the latter \citep{Molgedey1994SeparationCorrelations, Harrigan2017LandmarkDynamics}. SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data. This is a key attribute to our system as 10 million frames with 55 features in each frame are used as input data. The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}. For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.

@@

@@

In our coarse-grained SRV-MSM, we observe an "aligned hybridized" state, dissociated state, and four "shifted" states characterized by different combinations of out-of-register base pairings. These shifted states consist of 2 or 4 base pair shifting (single or doubled shifted) in either the 5' or 3' direction with varied stability and transition fluxes between kinetically neighboring states. This state decomposition is expected given the repetitive nature of the AT-all sequence and previous computational results that find shifted conformations form "deep kinetic traps" \citep{Xiao2019, Phys2014}. We found a substantial difference in thermodynamic stability and kinetic behavior between the 5'A shifted states and 3'T shifted states. This difference can be accounted for by examining experimental studies on the thermodynamics of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}. Dangling 5' ends have a consistently stabilizing effect, and inert tails tend to decrease stability as they increase in length, particularly at lower ionic strengths. It has been shown that 5' dangling ends with one inert tail have higher melting temperatures and are energetically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs}. This is might be attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturb the duplex structure and solvation environment \citep{Doktycz1990ThermodynamicATGC}. Furthermore, DNA nearest neighbor (NN) calculations that include dangling end contributions (Figure \ref{fig:AT-all_in_one}) predict that the specific nearest neighbors bonds in the 5' shifted state are more energetically favorable than those in the 3' shifted state. Although NN calculations were accurate in predicting the intact thermodynamics behavior of these sequences, they assume all-or-nothing hybridization and do not take inert tails into account \citep{Santalucia2004TM, SantaLucia1998AThermodynamics} Taken together, these experimental results substantiate why we observe the single-shifted 5' state to be only slightly less stable relative to the aligned hybridized state. Furthermore, we expect implied timescales of the two leading modes to be on the same order of magnitude as these reflect the average "dwell" time between transition into and out of these states. The stability of the single-shifted 3' state is notably lower than that of the single-shifted 5' state. In fact, the free energy is about equal to that of the double-shifted 5' state, despite it having two less intact base pairs. This is partially explained by nearest neighbor calculations, however these do not account for the differential effects of the 5' and 3' inert tails or kinetic considerations which we will discuss more below. The double-shifted 3' state is more stable than one might expect given the difference in stability between the single and double 5' shifted states. This might be attributed to the asymptotic effects of inert tail destabilization -- longer tails tend to have a diminishing effect on overall strand stability \citep{Michele2014EHybridization}. Accordingly, the longer inert tails in the double-shifted states have less of an effect on relative 5' vs. 3' stability compared to the shorter tails in the single-shifted states.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_in_one.PNG}
        \caption{a) Thermo value calculated from Santalucia data and from MSMs b) TICA projections of freen energies and state clustering c) and transitions probabilities between states}

@@

Beyond state probabilities and free energy approximations, the macrostate MSM yields valuable discrete kinetic information in the form of transition probabilities between states. Figure \ref{fig:AT-all_in_one} shows the probability of moving from one state to another within the MSM lag time (1.2 ns). All transition probabilities are higher when moving towards a more aligned state than towards a more shifted state, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability from double-shifted 3' state to the dissociated state is 6x higher than to the neighboring single-shifted 3' state. Although there is an approximately equal probability of transitioning from the dissociated state to the hybridized and 5' shifted states, there is a 2x lower probability of moving to the 3' shifted states. These asymmetric kinetic effects may have some role in the substantially lower state populations we see above. From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state. We recognize that these effects could be sensitive to the specific ion environment -- especially considering the effect of ionic strength on inert tails -- and warrant further computational and experimental investigation beyond this work.

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/skeleton/shifting_distribution.PNG}
        \caption{Shows similarities between shifting modes and state memberships for AT-all and GC-ends (Not sure if this adds anything, if anything can grey out underlying free energy landscape and move this to the SI.}
        \label{fig:shifting_distributions}
	\end{center}
\end{figure}

>>>>

\newcommand*{\rood}[1]{#1}
\newcommand*{\blauw}[1]{#1}
\newcommand*{\groen}[1]{#1}
\newcommand*{\blauwr}[1]{#1}
\newcommand{\Expect}[1]{\mathrm{E}\left[#1\right]}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

@@

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine}. Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales, and dynamic DNA nanotechnology has been used to perform basic calculation and to probe single molecules via temporal DNA signatures\citep{MhatreV.HoJi-AnnLee2012NIHAccess, Bui2018, Shah2019}. Both technologies rely on the hybridization reaction between complementary DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions. Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble meta-stable states facilitates the transition. Recent breakthrough studies have coupled experimental techniques with machine learning and MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}. Where these studies focus on association and dissociation kinetics alone, we broaden our analysis into higher order dynamical processes and meta-stable intermediates. The stability of certain intermediates states, such as out-of-register or shifted base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014, Xiao2019, Maciejczyk2014DNAModel}). Furthermore, frayed structures and dynamics have been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}. Sanstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability of frayed states was dictated by G:C base placement 10-mer oligonucleotide sequences \citep{Sanstead2016}. In this work, we study the same four sequences explored by Sanstead et al in an effort to uncover the sequence-dependent dynamics and their relation to metastable structures mentioned above.

Our understanding of DNA dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, or other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these present technical difficulties and are hampered by data collection rates \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices}. Förster resonance energy transfer (FRET) analysis, particularly when coupled with methods mentioned above, provides additional resolution, but it is unclear how fluorescent tags may interfere with the dynamics of short nucleotides \citep{Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several high-level computational models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}. Although these models provide experimentally verified speed-ups compared to all atom simulations, the long timescales on which DNA hybridization and dissociation events occur make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}.Other computational works use dramatically elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied certain aspect the overall dynamics process in one direction.

In this work, we use the coarse-grain 3 Sites Per Nucleotide (3spn.2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}. We perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another. Furthermore, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization  \citep{Jin2019, Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}. We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}. Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstates clustering can be performed. We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).

We find that GC base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior. By evaluating equilibrium trajectories we can study the relevance of meta-stable states during both the hybridization and dissociation process. Because SRVs generate an optimized low dimensional basis, we show that we can access higher resolution MSMs (about a 50\%reduction in required lag time) and generate more detailed models. Additionally, we can compare slow dynamical modes and meta-stable states between sequence-specific SRV-MSMs. Within these meta-stable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes. Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant DNA dynamics. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, meta-stable structures, and relative timescales.

@@

\subsection{\label{sec:methods}3spn.2 Model}

The 3 site per nucleotide (3spn.2) coarse grain model was designed to accelerate DNA computation relative to all atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}. Interactions sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing. The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}. 3spn.2 has been implemented to study DNA pack in viral capsids, protein-DNA binding, and nucleosome unwrapping, and recently served as a basis for the coarser 1cpn model \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping, Lequieu20191CPN:Chromatin}. The model is amenable to an implicit or explicit ion environment -- we chose the latter to include Mg2+ effects from experiment. We achieved better sampling of hybridization events with explicit ions, despite a more demanding computational cost.

@@

We initialized four sequences previously investigated by Sanstead et al. -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands according to 3SPN.2 documentation \citep{Sanstead2016, Phys2014}. We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}. In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end-to-end extension length of a single strand and the force cutoff (using Ewald summation method is set at 20 A). This translated to a box size of 77.74 Å and an effective oligo concentration of 7 mM. We used an Ewald potential to calculate long range Coulombic interaction between DNA and ions. We used a Debye-Huckel screening potential to account for phosphate-backbone interactions and preserve the persistence length and intrinsic curvature of DNA while taking the effects of ion-DNA interactions\citep{Hinckley2015}. We ran our simulations in the NVT ensemble and fix temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}. We set the simulation about 1 K higher than empirical sequence-specific melting temperature in order to maximize transitions between dissociated and hybridized states. We use a 15 fs time step and $3*10^{8}$ steps in each simulation, equating to 4.5$\mu$s simulation time. Base pair coordinates were saved every 30 ps, and full frames including backbones and ions were saved every 3 ns for visualization purposes. For each sequences, 100 simulations were performed in parallel, consuming about 32 serial cpu-hours of computation time for each independent simulation. An approximate Tmelt Boltzman distribution was replicated by initializing half of runs from the hybridized state and half from a random dissociated state. In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.

@@

All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}. Based on the self-complementary nature of each sequence -- meaning that both the sense and anti-sense strands are identical to each other -- we averaged permutable distances (45 pairs in total) together. This permutation reduction follows a similar procedure used in TICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}. The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}. The VAMP-2 score uses the covariances of a set of inputs estimates the transfer operator of a dynamical system (equation below), providing a robust and object means to evaluate various parameters and models. In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$. The VAMP-2 score can then be found for $\chi$ by applying the VAMP principle with cross-validation.

\begin{align*}
 	\mathscr{C}_{00}=&\Expect{\chi(x_t)\chi(x_t)^\intercal}_t\\
 	\mathscr{C}_{01}=&\Expect{\chi(x_t)\chi(x_{t+\tau})^\intercal}_t\\
 	\mathscr{C}_{11}=&\Expect{\chi(x_{t+\tau})\chi(x_{t+\tau})^\intercal}_{t+\tau}\\
	\\
 	VAMP2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
 	%VAMP2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
\end{align*}\label{CK1}

@@

SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The framework uses a twin-lobed artificial neural network, similar in structure to the network employed in VAMPnets \citep{Mardt2018VAMPnetsKinetics}, to learn a low dimensional representation of input features. This representation is optimized for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated. The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs. Other methods, such as time-lagged independent component analysis (TICA) and kernel TICA (kTICA), have been employed in place of SRVs but suffer limitations from the linear transformation in the case of the former and high computational cost in the case of the latter \citep{Molgedey1994SeparationCorrelations, Harrigan2017LandmarkDynamics}. SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data. This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence. The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}. For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.

@@

%Throughout this manuscript, we adopt "slithering" from previous work to describe dynamics between out-of-register states. We describe the metastable states along these pathway as "shifted."

@@

In our coarse-grained SRV-MSM, we observe an "aligned hybridized" state, dissociated state, and four "shifted" states characterized by different combinations of out-of-register base pairings. These shifted states consist of 2 or 4 base pair shifting (single or doubled shifted) in either the 5' or 3' direction with varied stability and transition fluxes between kinetically neighboring states. This state decomposition is expected given the repetitive nature of the AT-all sequence and previous computational results that find shifted conformations form "deep kinetic traps" \citep{Xiao2019, Phys2014}. We found a substantial difference in thermodynamic stability and kinetic behavior between the 5'A shifted states and 3'T shifted states. This difference can be accounted for by examining experimental studies on the thermodynamics of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}. Dangling 5' ends have a consistently stabilizing effect, and inert tails tend to decrease stability as they increase in length, particularly at lower ionic strengths. It has been shown that 5' dangling ends with one inert tail have higher melting temperatures and are energetically favorable compared to 3' ends \citep{Senior1988InfluenceDuplexes, Dickman2012ThermodynamicDNAs}. These effects might be attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturb the duplex structure and solvation environment \citep{Doktycz1990ThermodynamicATGC}. Furthermore, DNA nearest neighbor (NN) calculations that include dangling end contributions (Figure \ref{fig:AT-all_in_one}) predict that conformations in the 5' shifted state are more energetically favorable than those in the 3' shifted state. Although NN calculations were accurate in predicting the intact thermodynamics behavior of these sequences, they assume all-or-nothing hybridization and do not take inert tails into account \citep{Santalucia2004TM, SantaLucia1998AThermodynamics}. Therefore it makes sense that our PCCA+ free energies are all higher than thermodynamics prediction, due to some destabilizing effect of the inert tails. Notably, the 3' dangling states have a 6-8 kJ/mol higher free energy compared to theory, whereas the 5' dangling states deviate by less than 3 kJ/mol. In fact, the free energy of the 5'-double shifted state is slightly lower than that of the 3'-single shifted state, despite the former having two fewer intact base pairs. The double-shifted 3' state is more stable than one might expect given the difference in stability between the single and double 5' shifted states. This might be attributed to the asymptotic effects of inert tail destabilization -- longer tails tend to have a diminishing effect on overall strand stability \citep{Michele2014EHybridization}.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_0804/AT-all_in_one.png}
        \caption{a) AT-all free energies calculated from nearest neighbor data and from MSM PCCA+ probabilities b) TICA projections of free energies and PCCA+ state clusterings c) Transitions probabilities between states within the 1.2 ns lag time}

@@

Beyond state probabilities and free energy approximations, the macrostate MSM yields valuable discrete kinetic information in the form of transition probabilities between states. Figure \ref{fig:AT-all_in_one} shows the probability of moving from one state to another within the MSM lag time (1.2 ns). All transition probabilities are higher when moving towards a more aligned state than towards a more shifted state, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability from double-shifted 3' state to the dissociated state is 6x higher than to the neighboring single-shifted 3' state. Although there is an approximately equal probability of transitioning from the dissociated state to the hybridized and 5' shifted states, there is a 2x lower probability of moving to the 3' shifted states. These asymmetric kinetic effects, in addition to NN and inert tail contributions described above, may further lower the 3' shifted states populations relative to 5'. From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state. We recognize that these effects could be sensitive to the specific ion environment -- especially considering the effect of ionic strength on inert tails -- and warrant further computational and experimental investigation beyond this work.

@@
